Galois Characterization of Function Fields of \\ Transcendence Degree $1$ over Number Fields

### 2017-06-28 Master Thesis/synopsis.tex
Better phrasing

We denote the valuation ring of $v\in\operatorname{Spv}(K)$ by $(\mathcal{O}_v,\mathfrak{m}_v)$ and set $U_v=\mathcal{O}_v^\times$ and let $k(v)=\mathcal{O}_v/\mathfrak{m}_v$ be its residue field. The \textit{rank} of $v$ is defined as the Krull dimension of its valuation ring $\mathcal{O}_v$. The \textit{rational rank}\index{valuation!rational rank} of $v:K\to\Gamma_v\cup\{\infty\}$ is defined as the rank of its value group $\Gamma_v$ as a $\mathbb{Z}$-module:
\[\operatorname{rrank}(v) = \dim_\mathbb{Q}(\Gamma_v\otimes_\mathbb{Z}\mathbb{Q}) \]
\end{definition}

@@

\begin{definition}
We can endow $\operatorname{Spv}(K)$ with a natural topology by declaring following sets as a basis of open sets:
\[\{v\in\operatorname{Spv}(K)\mid v(x)\geq 0\text{ for all }x\in S\} \]
where $S$ ranges over all finite subsets of $K$. 
\end{definition}

@@

Let $K$ be a (not necessarily finitely generated) field over $\mathbb{Q}$ and of dimension $2$. First, let us recall some facts from Galois cohomology:

@@

@@

Now we calculate the cohomology groups $\mathrm{H}^3(K_w,\mathbb{Z}/\ell\mathbb{Z})$ for $w\in |\operatorname{Spv}(K)|$. With the help of Lemma \ref{2.4} we can find a cohomological criterion of henselian fields over $K$ with respect to a $w\in\operatorname{Spv}_\text{d}^{(2)}(K)$.

@@

Let $G_K(\ell)$ be the maximal pro-$\ell$ quotient of $G_K$. Since $\mu_\ell\subset K$, the action of $G_k$ on the $\ell$-Sylow group of the inertia group of $w$ factors through the maximal pro-$\ell$ quotient $G_k(\ell)$ of $G_k$. Since $G_{k,\ell}\cong G_k(\ell)$, we also see that the canonical projection $G_K\to G_K(\ell)$ restricts to an isomorphism $G_\ell\to G_K(\ell)$. Now consider the following composition:
\[ \begin{tikzcd}
\mathrm{H}^3(G_K(\ell),\mathbb{Z}/\ell\mathbb{Z}) \ar[r, "\operatorname{inf}"] & \mathrm{H}^3(G_K,\mathbb{Z}/\ell\mathbb{Z}) \ar[r, "\operatorname{res}"] & \mathrm{H}^3(G_\ell,\mathbb{Z}/\ell\mathbb{Z})
\end{tikzcd} \]
Since $\operatorname{res}\circ\operatorname{inf}$ is an isomorphism, $\operatorname{inf}$ is injective and $\operatorname{res}$ is surjective. We have $\operatorname{cor}\circ\operatorname{res} = [G_K:G_\ell]$. However, $[G_K:G_\ell]$ is coprime to $\ell$, which makes $\operatorname{res}$ injective as well. Therefore, $\operatorname{inf}$ and $\operatorname{res}$ are both isomorphisms and it suffices to calculate $\mathrm{H}^3(G_\ell,\mathbb{Z}/\ell\mathbb{Z})$. By \cite{Se64}, Ch. I, §4, Proposition 22 we have:
\[\operatorname{cd}(G_\ell) = \operatorname{cd}(\mathbb{Z}_\ell^r\rtimes G_k(\ell)) = \begin{cases}
r, & G_k(\ell) = 0 \\
r + 1, & G_k(\ell)\cong\mathbb{Z}_\ell
\end{cases} \] 
since $\operatorname{cd}(\mathbb{Z}_\ell)=1$. Thus $\mathrm{H}^3(G_\ell, \mathbb{Z}/\ell\mathbb{Z}) =0$ iff $G_k(\ell) = 0$ or $r < 2$. Now let $G_k(\ell) \cong\mathbb{Z}_\ell$ and $r = 2$. According to the proof of \cite{Se64}, Ch. I, §4, Proposition 22, we have:
\[ \mathrm{H}^3(G_\ell,\mathbb{Z}/\ell\mathbb{Z}) = \mathrm{H}^1(G_k(\ell), \mathrm{H}^2 (\mathbb{Z}_\ell^r, \mathbb{Z}/\ell\mathbb{Z})) \]
By [NSW] Theorem 2.4.6, we have $\mathrm{H}^2(\mathbb{Z}_\ell^r,\mathbb{Z}/\ell\mathbb{Z}) \cong\mathrm{H}^1(\mathbb{Z}_\ell, \mathrm{H}^1(\mathbb{Z}_\ell, \mathbb{Z}/\ell\mathbb{Z})) = \mathbb{Z}/\ell\mathbb{Z}$, since $\mathrm{H}^1(\mathbb{Z}_\ell, \mathbb{Z}/\ell\mathbb{Z})=\operatorname{Hom}_\text{cont.}(\mathbb{Z}_\ell, \mathbb{Z}/\ell\mathbb{Z})=\mathbb{Z}/\ell\mathbb{Z}$. Therefore, $\mathrm{H}^3(G_\ell,\mathbb{Z}/\ell\mathbb{Z})=\mathbb{Z}/\ell\mathbb{Z}$.

>>>>

We denote the valuation ring of $v\in\operatorname{Spv}(K)$ by $(\mathcal{O}_v,\mathfrak{m}_v)$, let $U_v=\mathcal{O}_v^\times$ be the unit group of $v$, and $k(v)=\mathcal{O}_v/\mathfrak{m}_v$ be its residue field. The \textit{rank} of $v$ is defined as the Krull dimension of its valuation ring $\mathcal{O}_v$. The \textit{rational rank}\index{valuation!rational rank} of $v:K\to\Gamma_v\cup\{\infty\}$ is defined as the rank of its value group $\Gamma_v$ as a $\mathbb{Z}$-module:
\[\operatorname{rrank}(v) = \dim_\mathbb{Q}(\Gamma_v\otimes_\mathbb{Z}\mathbb{Q}) \]
\end{definition}

@@

\begin{definition}
\[\{v\in\operatorname{Spv}(K)\mid v(x)\geq 0\text{ for all }x\in S\} \]
We can endow $\operatorname{Spv}(K)$ with a natural topology by declaring the following sets as a basis of open sets:
where $S$ ranges over all finite subsets of $K$. 
\end{definition}

@@

First, let us recall some facts from Galois cohomology:

@@

\paragraph{} Now let $K$ be a (not necessarily finitely generated) field over $\mathbb{Q}$ of dimension $2$. In this section, we want to give a cohomological characterization of $K$ being henselian.

@@

Now we calculate the cohomology groups $\mathrm{H}^3(K_w,\mathbb{Z}/\ell\mathbb{Z})$ for $w\in |\operatorname{Spv}(K)|$. With the help of our Main Lemma \ref{2.4} we can find a cohomological criterion of henselian fields over $K$ with respect to a $w\in |\operatorname{Spv}(K)|$.

@@

Let $G_K(\ell)$ be the maximal pro-$\ell$ quotient of $G_K$. Since $\mu_\ell\subset K$, the action of $G_k$ on the $\ell$-Sylow group of the inertia group of $w$ factors through the maximal pro-$\ell$ quotient $G_k(\ell)$ of $G_k$. Since $G_{k,\ell}\cong G_k(\ell)$, we also see that the canonical projection $G_K\to G_K(\ell)$ restricts to an isomorphism $G_\ell\to G_K(\ell)$. Now consider the following composition:
\[ \begin{tikzcd}
\mathrm{H}^3(G_K(\ell),\mathbb{Z}/\ell\mathbb{Z}) \ar[r, "\operatorname{inf}"] & \mathrm{H}^3(G_K,\mathbb{Z}/\ell\mathbb{Z}) \ar[r, "\operatorname{res}"] & \mathrm{H}^3(G_\ell,\mathbb{Z}/\ell\mathbb{Z})
\end{tikzcd} \]
Since $\operatorname{res}\circ\operatorname{inf}$ is an isomorphism, $\operatorname{inf}$ is injective and $\operatorname{res}$ is surjective. We have $\operatorname{cor}\circ\operatorname{res} = (G_K:G_\ell)$. However, $(G_K:G_\ell)$ is coprime to $\ell$, which makes $\operatorname{res}$ injective as well. Therefore, $\operatorname{inf}$ and $\operatorname{res}$ are both isomorphisms and it suffices to calculate $\mathrm{H}^3(G_\ell,\mathbb{Z}/\ell\mathbb{Z})$. By \cite{Se64}, Ch. I, §4, Proposition 22 we have:
\[\operatorname{cd}(G_\ell) = \operatorname{cd}(\mathbb{Z}_\ell^r\rtimes G_k(\ell)) = \begin{cases}
r, & G_k(\ell) = 0 \\
r + 1, & G_k(\ell)\cong\mathbb{Z}_\ell
\end{cases} \] 
since $\operatorname{cd}(\mathbb{Z}_\ell)=1$. Thus $\mathrm{H}^3(G_\ell, \mathbb{Z}/\ell\mathbb{Z}) =0$ iff $G_k(\ell) = 0$ or $r < 2$. Now let $G_k(\ell) \cong\mathbb{Z}_\ell$ and $r = 2$. According to the proof of \cite{Se64}, Ch. I, §4, Proposition 22, we have:
\[ \mathrm{H}^3(G_\ell,\mathbb{Z}/\ell\mathbb{Z}) = \mathrm{H}^1(G_k(\ell), \mathrm{H}^2 (\mathbb{Z}_\ell^r, \mathbb{Z}/\ell\mathbb{Z})) \]
By [NSW] Theorem 2.4.6, we have $\mathrm{H}^2(\mathbb{Z}_\ell^r,\mathbb{Z}/\ell\mathbb{Z}) \cong\mathrm{H}^1(\mathbb{Z}_\ell, \mathrm{H}^1(\mathbb{Z}_\ell, \mathbb{Z}/\ell\mathbb{Z})) = \mathbb{Z}/\ell\mathbb{Z}$, since $\mathrm{H}^1(\mathbb{Z}_\ell, \mathbb{Z}/\ell\mathbb{Z})=\operatorname{Hom}_\text{cont.}(\mathbb{Z}_\ell, \mathbb{Z}/\ell\mathbb{Z})=\mathbb{Z}/\ell\mathbb{Z}$. Therefore, $\mathrm{H}^3(G_\ell,\mathbb{Z}/\ell\mathbb{Z})=\mathbb{Z}/\ell\mathbb{Z}$.
